Efficient UQ for chemical systems using active \author{Manav Vohra, Alen Alexanderian, Hayley Guy, Sankaran Mahadevan} subspaces

### 2018-09-07 Paper/app.tex

In Figure~\ref{fig:hd} (left), we illustrate the resulting eigenvalue spectrum. The second eigenvalue is
found to be roughly two orders of magnitude smaller than the first, indicating that the active subspace
is 1-dimensional. This is further confirmed by the SSP plots in Figure~\ref{fig:hd} (right), generated
using both, perturbation-based and regression-based strategies. 
%
\begin{figure}[htbp]
 \begin{center}
   \includegraphics[width=0.45\textwidth]{./Figures/eig_33D}
   \includegraphics[width=0.45\textwidth]{./Figures/ssp_33D}
\caption{Left: Eigenvalue spectrum of $\hat{\mat{C}}$. Right: SSP of the computed active subspace, and the
corresponding 1-D linear fit.} 
\label{fig:hd}
\end{center}
\end{figure}
%
Specifically
in the case of perturbation-based strategy, the SSP is observed to exhibit a linear variation that
is reasonably
captured by the corresponding straight-line fit, $\tilde{G}$, constructed using the sequence of steps
outlined in Section~\ref{sub:ac}.
On the other hand, the SSP constructed using the regression-based strategy
is relatively more scattered. The two surrogates (straight-line fits) are observed to exhibit a
discrepancy unlike our earlier observation for the 19-dimensional problem in Figure~\ref{fig:comp_ssp}.

@@

The 1-dimensional surrogate ($\tilde{G}$) is investigated for its accuracy as well as the ability to capture the 
uncertainty in the
model output in two ways. Firstly, we estimate the relative L-2 norm of the discrepancy~($\varepsilon_d$)
between estimates of ignition delay in the case of H$_2$/O$_2$ reaction, obtained using the 
model output and the surrogate in the following equation:
%
\be
\varepsilon_d = \frac{\|G(\bm{\xi}) - \tilde{G}(\bm{\xi})\|_2}{\|G(\bm{\xi})\|_2}
\ee
%
Model evaluations ($G$) at 10$^{4}$ samples in the input domain were used to evaluate $\varepsilon_d$. 
It was estimated to be 1.35$\times10^{-2}$ and 1.02$\times10^{-1}$
using $\tilde{G}$ from the gradient-based approach and the gradient-free approach respectively. Hence,
an order of magnitude increase in $\varepsilon_d$ was estimated in the case of gradient-free approach.
Hence, in the norm-sense, it can be said that the 1-dimensional surrogate in the active subspace, obtained
using the gradient-based approach is remarkably accurate with a relative error of less than 1.5$\%$, whereas
the gradient-free approach resulted in a relative error of about 10$\%$.

Secondly, we verify the accuracy of the two surrogates in a probabilistic setting. In particular, we compare 
probability density functions (PDFs) obtained using the true set of model evaluations, and 1-dimensional
surrogates from the two approaches, in Figure~\ref{fig:pdf_33D}. Note that the three PDFs were evaluated 
using the same set of 10$^4$ samples in the cross-validation set. 
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=3.0in]{./Figures/pdf_comp_id_1e4}
\end{center} 
\caption{A comparison of the PDFs of ignition delay, obtained using model 
evaluations (solid line) and 1-dimensional surrogates using the gradient-free approach (dashed line) and the gradient-based
approach (dashed-dotted line). The same set of 10$^4$ samples in the cross-validation set were used in each case.}
\label{fig:pdf_33D}
\end{figure}
%
%
%\begin{figure}[htbp]
% \begin{center}
%  \includegraphics[width=0.45\textwidth]{./Figures/pdf_comp_id_1e4}
%\caption{A comparison of the PDFs of ignition delay, obtained using model evaluations (solid line) and
%1-dimensional surrogates using the gradient-free approach (dashed line) and the gradient-based approach
%(dashed-dotted line). The same set of 10$^4$ samples in the cross-validation set were used in each case.}
%\label{fig:pdf_33D}
%\end{center}
%\end{figure}
%
The PDFs based on the model evaluations and the gradient-based approach are nearly identical, confirming
the existence of a 1-dimensional active subspace as well as the accuracy of the gradient-based approach. 
However, though the PDF based on the gradient-free approach is observed to show consistency with the true 
PDF in the modal
estimate; it underestimates the spread in the ignition delay. Specific values of the mean and the standard deviation
in each case are provided in Table~\ref{tab:stats}. 
%
\begin{table}[htbp]
\begin{center}
\begin{tabular}{ccc}
\toprule
$\textbf{Distribution}$ & $\mu$ & $\sigma$ \\ 
\bottomrule
$G$~(Model) & 0.133 & 0.0198 \\
$\tilde{G}$~(Gradient-based) & 0.133 & 0.0196 \\
$\tilde{G}$~(Gradient-free) & 0.133 & 0.0167 \\
\bottomrule
\end{tabular}
\caption{The mean ($\mu$), and the standard deviation ($\sigma$), computed using the model ($G$), and
the surrogate ($\tilde{G}$) based on the two approaches at 10$^4$ samples in the cross-validation
set.}
\label{tab:stats}
\end{center}
\end{table}
%
The three PDFs are found to be
consistent in their prediction of mean value of the ignition delay.  The standard deviation estimates based on model
evaluations and the gradient-based approach are found to be in close agreement, whereas, it  is accurately estimated
by the gradient-free approach only upto the first significant figure.

>>>>

In Figure~\ref{fig:hd} (left), we illustrate the resulting eigenvalue spectrum. The second eigenvalue is
found to be roughly two orders of magnitude smaller than the first, indicating that the active subspace
is 1-dimensional. This is further confirmed by the SSP plots in Figure~\ref{fig:hd} (right), generated
using both, perturbation-based and regression-based strategies. 
%
\begin{figure}[htbp]
 \begin{center}
   \includegraphics[width=0.45\textwidth]{./Figures/eig_33D}
   \includegraphics[width=0.45\textwidth]{./Figures/ssp_33D}
\caption{Left: Eigenvalue spectrum of $\hat{\mat{C}}$. Right: SSP of the computed active subspace, and the
corresponding 1-D linear fit.} 
\label{fig:hd}
\end{center}
\end{figure}
%
Specifically
in the case of perturbation-based strategy, the SSP is observed to exhibit a linear variation that
is reasonably
captured by the corresponding straight-line fit, $\tilde{G}$, constructed using the sequence of steps
outlined in Section~\ref{sub:ac}.
On the other hand, the SSP constructed using the regression-based strategy
is relatively more scattered. Consequently, the two surrogates (straight-line fits) are observed to exhibit a
discrepancy unlike our earlier observation for the 19-dimensional problem in Figure~\ref{fig:comp_ssp}.

@@

The 1-dimensional surrogate ($\tilde{G}$) shown in Figure~\ref{fig:hd}~(right) is investigated
for its accuracy as well as the ability to capture the uncertainty in the
model output in two ways. Firstly, we estimate the relative L-2 norm of the discrepancy~($\varepsilon_d$)
between estimates of ignition delay in the case of H$_2$/O$_2$ reaction, obtained using the 
model output and the surrogate in the following equation:
%
\be
\varepsilon_d = \frac{\|G(\bm{\xi}) - \tilde{G}(\bm{\xi})\|_2}{\|G(\bm{\xi})\|_2}
\ee
%
Model evaluations ($G$) at 10$^{4}$ samples in the input domain were used to evaluate $\varepsilon_d$. 
It was estimated to be 1.35$\times10^{-2}$ and 1.02$\times10^{-1}$
using $\tilde{G}$ based on the perturbation-based and the regression-based strategies respectively. Hence,
$\varepsilon_d$ was found to increase by an order of magnitude when using the regression-based strategy.
Therefore, in the norm-sense, it can be said that the 1-dimensional surrogate in the active subspace, obtained
using the perturbation-based approach is remarkably accurate with a relative error of less than 1.5$\%$, whereas
the regression-based approach resulted in a relative error of about 10$\%$.

Secondly, we verified the accuracy of the two surrogates in a probabilistic setting. In particular, we compared 
probability density functions (PDFs) obtained using the true set of model evaluations, and 1-dimensional
surrogates ($\tilde{G}$'s) from the two strategies, in Figure~\ref{fig:pdf_33D}. Note that the three PDFs were evaluated 
using the same set of 10$^4$ samples in the cross-validation set. 
%
\begin{figure}[htbp]
\begin{center}
\includegraphics[width=3.0in]{./Figures/pdf_comp_id_1e4}
\end{center} 
\caption{A comparison of the PDFs of ignition delay, obtained using model 
evaluations (solid line) and 1-dimensional surrogates using the regression-based strategy (dashed line) and the
perturbation-based strategy (dashed-dotted line). The same set of 10$^4$ samples in the cross-validation set were 
used in each case.}
\label{fig:pdf_33D}
\end{figure}
%
%
%\begin{figure}[htbp]
% \begin{center}
%  \includegraphics[width=0.45\textwidth]{./Figures/pdf_comp_id_1e4}
%\caption{A comparison of the PDFs of ignition delay, obtained using model evaluations (solid line) and
%1-dimensional surrogates using the gradient-free approach (dashed line) and the gradient-based approach
%(dashed-dotted line). The same set of 10$^4$ samples in the cross-validation set were used in each case.}
%\label{fig:pdf_33D}
%\end{center}
%\end{figure}
%
The PDFs based on the model evaluations and the perturbation-based strategy are nearly identical, confirming
the existence of a 1-dimensional active subspace as well as the accuracy of this strategy. 
On the other hand, although the PDF obtained using the regression-based strategy is observed to show
consistency with the true PDF in the modal
estimate, it clearly underestimates the spread in the ignition delay. Specific values of the mean and the standard 
deviation in each case are provided in Table~\ref{tab:stats}. 
%
\begin{table}[htbp]
\begin{center}
\begin{tabular}{ccc}
\toprule
$\textbf{Distribution}$ & $\mu$ & $\sigma$ \\ 
\bottomrule
$G$~(Model) & 0.133 & 0.0198 \\
$\tilde{G}$~(Perturbation-based) & 0.133 & 0.0196 \\
$\tilde{G}$~(Regression-based) & 0.133 & 0.0167 \\
\bottomrule
\end{tabular}
\caption{The mean ($\mu$), and the standard deviation ($\sigma$), computed using the model ($G$), and
the surrogate ($\tilde{G}$) based on the two strategies at 10$^4$ samples in the cross-validation
set.}
\label{tab:stats}
\end{center}
\end{table}
%
The three PDFs are found to be
consistent in their prediction of mean value of the ignition delay.  The standard deviation estimates based on model
evaluations and the perturbation-based strategy are found to be in close agreement, whereas, it  is accurately
 estimated in the case of regression-based strategy only upto the first significant digit.

### 2018-09-07 Paper/intro.tex

We seek to understand the impact of uncertainty in the
rate-controlling parameters, pre-exponent~($A_i$) and the activation
energy~($E_{a,i}$) on the ignition delay. The $A_i$'s associated with all
reactions and the $E_{a,i}$'s with non-zero nominal estimates
are considered to be uniformly distributed about their nominal estimates provided
in~\cite{Yetter:1991}.  The total number of uncertain inputs is 33 which makes
the present problem computationally challenging due to the large number of 
uncertain parameters.  
To address this challenge, we focus on reducing the dimensionality
of the problem by computing the aforementioned active subspace.
This involves repeated evaluations of the gradient of a model output with
respect to the input parameters. Estimating the
gradients can however be computationally challenging in high-dimensional settings. 
Hence, in this work, we also explore the suitability of a gradient-free
approach that involves a regression-based local linear approximation of
the model output as discussed later in~\ref{sub:gradfree}. The two approaches
 are based on
algorithms 1.1 and 1.2 in~\cite{Constantine:2015}. Both strategies
are implemented in an iterative manner i.e. a new set of model evaluations are
generated, and gradient evaluations are performed at each iteration until 
a reasonably accurate approximation to the active subspace is achieved, as
discussed later in Section~\ref{sec:method}.

An alternate approach to dimension reduction involves computing the global
sensitivity measures associated with the uncertain inputs of a model. Depending
upon the estimates of the sensitivity measures, only the important inputs are
varied for the purpose of uncertainty quantification (UQ). Sobol' indices are
commonly used as global sensitivity measures~\cite{Sobol:2001}. They are
used to quantify the relative contributions of the uncertain inputs to the variance
in the output, either individually, or in combination with other inputs. 
Multiple efforts have focused on efficient computation of the Sobol' 
indices~\cite{Sudret:2008,Plischke:2013,Tissot:2015,Li:2016} including the 
derivative-based global sensitivity measures~(DGSMs), developed to
compute approximate upper bounds for the Sobol' indices with much fewer
computations~\cite{Sobol:2009, Lamboni:2013}. It was noted
in~\cite{Diaz:2016,Constantine:2017} that DGSMs can be approximated by
exploiting their links with active subspaces. This led to definition of the so
called \emph{activity scores}. In Section~\ref{sub:gsa}, we build on these
ideas to provide a complete analysis of links between Sobol indices, DGSMs, and
activity scores for functions of independent random inputs whose distribution
law belongs to the class of Boltzmann probability measures. 
It is worth mentioning that computing global sensitivity measures provides 
important information about a model that go beyond dimension reduction. By 
identifying parameters with significant impact on the model output, we can assess
regimes of validity of the model formulation, and gain critical insight into the
underlying physics in many cases.

The main contributions of this paper are as follows: 
\begin{itemize}
\item 
Active subspace
discovery in a high-dimensional H$_2$/O$_2$ kinetics problem involving 33
uncertain inputs: The methodology presented in this work
resulted in a 1-dimensional active subspace, indicating immense potential
for computational savings. The presented
analysis can also guide practitioners in other problems of chemical kinetics on using the
method of active subspaces to achieve efficiency in uncertainty propagation.  
\item Comprehensive numerical investigation of the gradient-based and gradient-free approach: 
We investigate the suitability of both approaches
for computing active subspaces, surrogate modeling, and approximating global
sensitivity measures through a comprehensive set of numerical experiments. Our
results reveal insight into the merits of the methods approach as well as
their shortcomings.  
\item Analysis of the links between  
global sensitivity measures: 
By connecting the recent theoretical advances in variance-based and
derivative-based global sensitivity analysis to active subspaces, we provide a complete analysis
of the links between total Sobol' indices, DGSMs, and activity scores for a broad
class of probability distributions. Our analysis is concluded by a result quantifying
approximation errors incurred due to fixing unimportant parameters, deemed so by 
computing their activity scores.

@@

This article is organized as follows. In section~\ref{sub:ac}, a brief
theoretical background on the active subspace methodology is provided. In
section~\ref{sub:gsa}, it is shown that the activity scores provide a
reasonable approximation to the DGSMs especially in a high-dimensional setting.
Additionally, a relationship between the three global sensitivity measures,
namely, the activity scores, DGSMs, and the total Sobol' indices is
established. In section~\ref{sec:method}, a systematic framework for computing
the active subspace using an iterative gradient-based approach is provided.
Numerical results based on the gradient-free approach are compared with those
obtained using the gradient-based approach.  The active subspace is computed
for the H$_2$/O$_2$ reaction kinetics problem whereby only the $A_i$'s are
considered as uncertain. In this case, the two approaches are observed to yield
consistent results. In section~\ref{sec:app}, the gradient-based approach is
further implemented to compute the active subspace for the higher-dimensional
H$_2$/O$_2$ reaction kinetics problem. The resulting active subspace is once again
found to be 1-dimensional, thereby indicating enormous potential for computational
savings in this application. The active subspace is also computed using the
gradient-free approach. It is observed that although the uncertainty in the
ignition delay is reasonably approximated, the gradient-free approach fails to
capture the global sensitivity of all the uncertain inputs. Hence, a trade-off
between efficiency and accuracy between the two approaches is observed in this
case.  Finally, a summary and discussion based on our findings is included in
section~\ref{sec:conc}.

>>>>

We seek to understand the impact of uncertainty in the
rate-controlling parameters, pre-exponent~($A_i$) and the activation
energy~($E_{a,i}$) on the ignition delay. The $A_i$'s associated with all
reactions and the $E_{a,i}$'s with non-zero nominal estimates
are considered to be uniformly distributed about their nominal estimates provided
in~\cite{Yetter:1991}.  The total number of uncertain inputs is 33 which makes
the present problem computationally challenging due to the large number of 
uncertain parameters.  
To address this challenge, we focus on reducing the dimensionality
of the problem by computing the aforementioned active subspace.
This involves repeated evaluations of the gradient of a model output with
respect to the input parameters. Several numerical techniques are available
for computing the gradient. Perturbation-based techniques
include finite differences and more advanced methods involving
adjoints~\cite{Jameson:1988,Gunzburger:2003,Borzi:2011,Alexanderian:2017}. The
adjoint-based method requires a solution of the state equation (forward solve)
and the corresponding adjoint equation. Hence, it is 
limited by the availability of an adjoint solver. Additional model evaluations
at neighboring points are required if finite difference is used which increases
the computational effort. Regression-based techniques aim to estimate the
gradient by approximating the model output using a regression fit. These
are computationally less intensive than the former. However, as expected,
there is a trade-off between computational effort and accuracy in the two
approaches for estimating the gradient.

In this work, we implement two strategies for computing the active subspace for the
H$_2$/O$_2$ mechanism, and refer to them as \textit{perturbation-based} and 
\textit{regression-based} to highlight the difference in the approach for estimating
the gradient of the model output as discussed.  
Other than the difference in gradient estimation, the two strategies follow the same
sequence of steps provided in~\cite{Constantine:2015}. However, 
in order to reduce computational effort, the two strategies are implemented in an
iterative manner in this work as discussed later in Section~\ref{sec:method}.

An alternate approach to dimension reduction involves computing the global
sensitivity measures associated with the uncertain inputs of a model. Depending
upon the estimates of the sensitivity measures, only the important inputs are
varied for the purpose of uncertainty quantification (UQ). Sobol' indices are
commonly used as global sensitivity measures~\cite{Sobol:2001}. They are
used to quantify the relative contributions of the uncertain inputs to the variance
in the output, either individually, or in combination with other inputs. 
Multiple efforts have focused on efficient computation of the Sobol' 
indices~\cite{Sudret:2008,Plischke:2013,Tissot:2015,Li:2016} including the 
derivative-based global sensitivity measures~(DGSMs), developed to
compute approximate upper bounds for the Sobol' indices with much fewer
computations~\cite{Sobol:2009, Lamboni:2013}. It was noted
in~\cite{Diaz:2016,Constantine:2017} that DGSMs can be approximated by
exploiting their links with active subspaces. This led to the definition of the 
so-called \emph{activity scores}. In Section~\ref{sub:gsa}, we build on these
ideas to provide a complete analysis of links between Sobol indices, DGSMs, and
activity scores for functions of independent random inputs whose distribution
law belongs to the class of Boltzmann probability measures. 
It is worth mentioning that computing global sensitivity measures provides 
important information about a model that go beyond dimension reduction. By 
identifying parameters with significant impact on the model output, we can assess
regimes of validity of the model formulation, and gain critical insight into the
underlying physics in many cases.

The main contributions of this paper are as follows: 
\begin{itemize}
\item 
Active subspace
discovery in a high-dimensional H$_2$/O$_2$ kinetics problem involving 33
uncertain inputs: The methodology presented in this work
resulted in a 1-dimensional active subspace, indicating immense potential
for computational savings. The presented
analysis can also guide practitioners in other problems of chemical kinetics on using the
method of active subspaces to achieve efficiency in uncertainty propagation.  
\item Comprehensive numerical investigation of the perturbation-based and the regression-based approaches: 
We investigate the suitability of both approaches
for computing active subspaces, surrogate modeling, and approximating global
sensitivity measures through a comprehensive set of numerical experiments. Our
results reveal insight into the merits of the methods as well as
their shortcomings.  
\item Analysis of the links between  
global sensitivity measures: 
By connecting the recent theoretical advances in variance-based and
derivative-based global sensitivity analysis to active subspaces, we provide a complete analysis
of the links between total Sobol' indices, DGSMs, and activity scores for a broad
class of probability distributions. Our analysis is concluded by a result quantifying
approximation errors incurred due to fixing unimportant parameters, deemed so by 
computing their activity scores.

@@

This article is organized as follows. In section~\ref{sub:ac}, a brief
theoretical background on the active subspace methodology is provided. In
section~\ref{sub:gsa}, it is shown that the activity scores provide a
reasonable approximation to the DGSMs especially in a high-dimensional setting.
Additionally, a relationship between the three global sensitivity measures,
namely, the activity scores, DGSMs, and the total Sobol' indices is
established. In section~\ref{sec:method}, a systematic framework for computing
the active subspace is provided. 
Numerical results based on the perturbation-based approach are compared with those
obtained using the regression-based approach.  For our initial investigation aimed
as assessing the feasibility and suitability of the two strategies, the active 
subspace is computed for a 19-dimensional 
H$_2$/O$_2$ reaction kinetics problem whereby only the $A_i$'s are
considered as uncertain. In this case, the two approaches are observed to yield
consistent results. Motivated by these results, we further implement the two
strategies to compute the active subspace for a 33-dimensional H$_2$/O$_2$
reaction kinetics problem in section~\ref{sec:app}. Once again, the two
strategies reveal the existence of a 1-dimensional active subspace. However,
it is observed that although the regression-based strategy results in a 
reasonable approximation of the mean, mode, and the uncertainty in the ignition
delay, it fails to capture the global sensitivity of all the uncertain inputs. 
Hence, a trade-off between efficiency and accuracy between the two approaches is
apparent in this case.  Finally, a summary and discussion based on our findings is included in
section~\ref{sec:conc}.
